---
title: "GPS_peng_mapping"
author: "Taylor Azizeh"
date: "2025-02-16"
output: html_document
---

############################################################################################################
*Title*: Plotting and animation for foraging emperor penguin GPS tracks
*Author*: Taylor Azizeh
*Date*: 29 May 2025
*Description*: This code takes GPS tracks (interpolated to five minute intervals) of foraging emperor penguins off Cape Crozier, Antarctica, and creates track plots, as well as animations that can be exported as .gifs for science communication or presentations. For this code, I downloaded and loaded in the International Bathymetric Chart of the Southern Ocean (IBSCO) bathymetry and sea ice rasters (https://ibcso.org/current_version/). For your study system, you're going to want to find a bathymetric raster.
*Contact*: taylorazizeh@gmail.com

############################################################################################################

## Step 1: Load and clean data.

```{r}
# Read in libraries
library(data.table) # for reading in data
library(sf) # for dealing with spatial objects
library(ggplot2) # for mapping
library(rnaturalearth) # for mapping
library(dplyr) # for moving data around
library(ggspatial)
library(terra)
library(transformr)
library(raster)
```

```{r}
# Read in GPS points
gps <- read.csv('/Users/taylorazizeh/Documents/research/active/Emperor penguins/data/cleaned/gps interpolated/fixed timestamps/cleaned_GPS_data_all_years.csv')
head(gps)
```

```{r}
# Create a spatial object from the GPS points
gps_sf <- st_as_sf(gps, coords = c("Lon","Lat"), crs = 4326)

# Check out imported data
summary(gps_sf)
st_crs(gps_sf)  # The CRS for these points is EPSG: 3031 or Antarctic Polar Stereographic
```

## Step 2: Prep and load in bathymetry and ice data.
``` {r}
# Load IBCSO bathymetry raster
ibcso_bedrock_path <- "/Users/taylorazizeh/Documents/research/active/Emperor penguins/data/cleaned/bathymetry/IBCSO_v2_bed_RGB.tif" 
ibcso_ice_path <- "/Users/taylorazizeh/Documents/research/active/Emperor penguins/data/cleaned/bathymetry/IBCSO_v2_ice-surface_RGB.tif"

# Convert to a raster
ibcso_bedrock <- rast(ibcso_bedrock_path)
ibcso_ice <- rast(ibcso_ice_path)

# Define region of interest
xmin <- -100000
xmax <- 500000
ymin <- -1550000

ymax <- -1250000
extent_of_interest <- ext(xmin, xmax, ymin, ymax)

# Crop bathymetry and ice rasters
ibcso_bedrock_cropped <- crop(ibcso_bedrock, extent_of_interest)
ibcso_ice_cropped <- crop(ibcso_ice, extent_of_interest)

# Convert bathymetry raster to data frame for ggplot
bedrock_df <- as.data.frame(ibcso_bedrock_cropped, xy = TRUE) %>%
  mutate(color = rgb(
    IBCSO_v2_bed_RGB_1 / 255,
    IBCSO_v2_bed_RGB_2 / 255,
    IBCSO_v2_bed_RGB_3 / 255
  ))

# Convert ice raster to data frame for ggplot
ice_df <- as.data.frame(ibcso_ice_cropped, xy = TRUE) %>%
  mutate(color = rgb(
    `IBCSO_v2_ice-surface_RGB_1` / 255,
    `IBCSO_v2_ice-surface_RGB_2` / 255,
    `IBCSO_v2_ice-surface_RGB_3` / 255
  ))
```

## Step 3: Format and clean GPS tracks and create plot.

```{r}
# Remove birds with none or little GPS data from dataset
gps_no10c_08c <- gps %>%
  filter(!(ID %in% c("22EP_310c", "22EP_308c")))

# Convert penguin data to sf object (WGS84)
pen_data_sf <- st_as_sf(
  gps_no10c_08c,
  coords = c("Lon", "Lat"),
  crs = 4326
)

# Transform penguin data to match IBCSO CRS
pen_data_sf <- st_transform(pen_data_sf, crs(ibcso_bedrock_cropped))

# Color blind friendly
cb_palette_no_blues <- c(
  "#000000",  # black
  "#E69F00",  # orange
  "#009E73",  # bluish green (ok over blue background)
  "#F0E442",  # yellow
  "#D55E00",  # vermilion
  "#CC79A7",  # reddish purple
  "#999999",  # gray
  "#A65628",  # brown
  "#FF7F00",  # vivid orange
  "#984EA3",  # purple
  "#4DAF4A",  # medium green
  "#F781BF",  # pink
  "#999933",  # olive
  "#E41A1C",  # red
  "#FC8D62",  # salmon
  "#66C2A5",  # sea green
  "#E78AC3",  # light magenta
  "#A6D854",  # lime green
  "#FFD92F",  # gold
  "#E5C494",  # beige
  "#B15928",  # dark brown
  "#FB9A99",  # light pink
  "#CAB2D6",  # lavender
  "#FFFF99",  # pale yellow
  "#B2DF8A"   # soft green
)

# Convert back to data frame for ggplot
pen_data_transformed <- pen_data_sf %>%
  st_coordinates() %>%
  as.data.frame() %>%
  bind_cols(pen_data_sf %>% st_drop_geometry()) %>%
  rename(x = X, y = Y)

# Plotting
fig1 <- ggplot() +
  
  # Add bathymetry raster
  geom_raster(
    data = bedrock_df,
    aes(x = x, y = y, fill = color),
    show.legend = FALSE
  ) +
  scale_fill_identity() +
  
  # Add ice raster with transparency
  geom_raster(
    data = ice_df,
    aes(x = x, y = y, fill = color),
    alpha = 0.7,
    show.legend = FALSE
  ) +
  
  # Add penguin tracks, color-coded by ID
  geom_point(
    data = pen_data_transformed,
    aes(x = x, y = y, color = ID),
    size = 0.3,
    alpha = 0.3
  ) +
  scale_color_manual(values = setNames(cb_palette_no_blues, sort(unique(pen_data_transformed$ID)))) +

  # Add scale bar and north arrow
  annotation_scale(location = "bl", width_hint = 0.2) +
  annotation_north_arrow(location = "tl",
                         style = north_arrow_orienteering(),
                         height = unit(1.0, "cm"),
                         width = unit(1.0, "cm"),
                         pad_x = unit(0.25, "cm"),
                         pad_y = unit(0.25, "cm")) +
  
  # Add titles and labels
  labs(
    x = "Longitude",
    y = "Latitude"
  ) +
  
  # Coordinate system and theme
  coord_sf(crs = st_crs(ibcso_bedrock_cropped), expand = FALSE) +
  theme_bw(base_size = 14) +
  theme(legend.text = element_text(size = 7),
        legend.title = element_text(size = 10),
        legend.key.width = unit(0.05, "lines"),
        axis.text = element_text(size = 6),
        axis.title = element_text(size = 8)) +
  guides(color = guide_legend(ncol = 2,
    override.aes = list(shape = 16, size = 3, alpha = 1)
  ))

print(fig1)

# Save Figure 1
# ggsave(fig1, 
#        filename = "RMap.jpeg",
#        device = "jpeg",
#        path = "/Users/taylorazizeh/Documents/research/active/Emperor penguins/figures/tracks",
#        height = 5, width = 7, units = "in")
```

## Step 4: Create gif of animal movement.

```{r}
# Ensure you retain timestamp & year info
gps_clean <- gps_no10c_08c %>%
  mutate(
    Timestamp = as.POSIXct(Timestamp),
    Year = year(Timestamp)
  )

# Transform to sf, then to projected coords
pen_data_sf <- st_as_sf(gps_clean, coords = c("Lon", "Lat"), crs = 4326)
pen_data_sf <- st_transform(pen_data_sf, crs(ibcso_bedrock_cropped))

# Keep coords + metadata
pen_data_transformed <- pen_data_sf %>%
  st_coordinates() %>%
  as.data.frame() %>%
  bind_cols(pen_data_sf %>% st_drop_geometry()) %>%
  rename(x = X, y = Y)

# Filter to 2019
pen_2019 <- pen_data_transformed %>% filter(Year == 2019)

# Set up IDs for plotting
pen_2019$ID <- factor(pen_2019$ID, levels = sort(unique(pen_2019$ID)))

# Plot: 2019 penguin tracks
fig_2019 <- ggplot() +
  geom_raster(data = bedrock_df, aes(x = -x, y = -y, fill = color), show.legend = FALSE) +
  scale_fill_identity() +
  geom_raster(data = ice_df, aes(x = -x, y = -y, fill = color), alpha = 0.7, show.legend = FALSE) +
  geom_path(data = pen_2019, aes(x = -x, y = -y, group = ID, color = ID), linewidth = 0.5, alpha = 0.8) +
  annotation_scale(location = "bl", width_hint = 0.2) +
  annotation_north_arrow(location = "tr", style = north_arrow_orienteering()) +
  labs(
    title = "2019 Penguin Tracks",
    x = "Projected X (meters)", y = "Projected Y (meters)"
  ) +
  coord_sf(crs = st_crs(ibcso_bedrock_cropped), expand = FALSE) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  guides(color = guide_legend(ncol = 2)) +
  scale_color_manual(values = setNames(cb_palette_no_blues, sort(unique(pen_2019$ID))))

# Save or print
# ggsave("figure_2019.jpeg", fig_2019, width = 7, height = 5, units = "in", dpi = 300)

anim_2019 <- ggplot() +
  geom_raster(data = bedrock_df, aes(x = -x, y = -y, fill = color), show.legend = FALSE) +
  scale_fill_identity() +
  geom_raster(data = ice_df, aes(x = -x, y = -y, fill = color), alpha = 0.7, show.legend = FALSE) +
  geom_path(
    data = pen_2019,
    aes(x = -x, y = -y, group = ID, color = ID),
    linewidth = 1,
    show.legend = FALSE
  ) +
  scale_color_manual(values = setNames(cb_palette_no_blues, sort(unique(pen_2019$ID)))) +
  coord_sf(crs = st_crs(ibcso_bedrock_cropped), expand = FALSE) +
  labs(
    title = "Emperor penguin movement\n{format(frame_time, '%b %d, %Y')}",
    x = "Latitude",
    y = "Longitude"
  ) +
  transition_time(Timestamp) +  # instead of transition_reveal
  ease_aes("linear") +
  theme_minimal(base_size = 14)

# Render the animation
anim_2019 <- gganimate::animate(anim, renderer = gifski_renderer(), fps = 10, width = 800, height = 600, nframes = 150)

# Preview
anim_2019

# Export
# anim_save(filename = "/Users/taylorazizeh/Documents/research/active/Emperor penguins/figures/penguin_tracks_2019.gif", animation = anim_2019)
```

```{r}
# Transform to sf, then to projected coords
pen_data_sf <- st_as_sf(gps_clean, coords = c("Lon", "Lat"), crs = 4326)
pen_data_sf <- st_transform(pen_data_sf, crs(ibcso_bedrock_cropped))

# Keep coords + metadata
pen_data_transformed <- pen_data_sf %>%
  st_coordinates() %>%
  as.data.frame() %>%
  bind_cols(pen_data_sf %>% st_drop_geometry()) %>%
  rename(x = X, y = Y)

# Filter to 2019
pen_2022 <- pen_data_transformed %>% filter(Year == 2022)

# Plot: 2019 penguin tracks
fig_2022 <- ggplot() +
  geom_raster(data = bedrock_df, aes(x = -x, y = -y, fill = color), show.legend = FALSE) +
  scale_fill_identity() +
  geom_raster(data = ice_df, aes(x = -x, y = -y, fill = color), alpha = 0.7, show.legend = FALSE) +
  geom_path(data = pen_2019, aes(x = -x, y = -y, group = ID, color = ID), linewidth = 0.5, alpha = 0.8) +
  annotation_scale(location = "bl", width_hint = 0.2) +
  annotation_north_arrow(location = "tr", style = north_arrow_orienteering()) +
  labs(
    title = "2022 Penguin Tracks",
    x = "Longitude", y = "Latitude"
  ) +
  coord_sf(crs = st_crs(ibcso_bedrock_cropped), expand = FALSE) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  guides(color = guide_legend(ncol = 2))

# Save or print
# ggsave("figure_2022.jpeg", fig_2022, width = 7, height = 5, units = "in", dpi = 300)

anim_2022  <- ggplot() +
  geom_raster(data = bedrock_df, aes(x = -x, y = -y, fill = color), show.legend = FALSE) +
  scale_fill_identity() +
  geom_raster(data = ice_df, aes(x = -x, y = -y, fill = color), alpha = 0.7, show.legend = FALSE) +
  geom_path(
    data = pen_2022,
    aes(x = -x, y = -y, group = ID, color = ID),
    linewidth = 1,
    show.legend = FALSE
  ) +
  scale_color_manual(values = setNames(cb_palette_no_blues, sort(unique(pen_2022$ID)))) +
  coord_sf(crs = st_crs(ibcso_bedrock_cropped), expand = FALSE) +
  labs(
    title = "Emperor penguin movement",
    x = "Latitude",
    y = "Longitude"
  ) +
  transition_reveal(Timestamp) +
  ease_aes("linear") +
  theme_minimal(base_size = 14)

# Render the animation
anim_2022 <- gganimate::animate(anim, renderer = gifski_renderer(), fps = 10, width = 800, height = 600, nframes = 150)

# Save optional
anim_save(filename = "/Users/taylorazizeh/Documents/research/active/Emperor penguins/figures/penguin_tracks_2022.gif", animation = anim_2022)
```


